Cross-Loop Optimization of Arithmetic Intensity for Finite Element Local Assembly

### 2014-09-10 TACO-paper-v2/coffee.tex
Some minor English fixes

In many fields such as computational fluid dynamics, computational
electromagnetics and structural mechanics, phenomena are modelled by
partial differential equations (PDEs). Numerical techniques, like the
finite volume method and the finite element method, are widely employed
to approximate solutions of these PDEs. Unstructured meshes are often
used to discretize the computational domain, since they allow an
accurate representation of complex geometries. The solution is sought by
applying suitable numerical operations, described by computational
kernels, to all entities of a mesh, such as edges, vertices, or cells.
On standard clusters of multicores, typically, a kernel is executed
sequentially by a thread, while parallelism is achieved by partitioning
the mesh and assigning each partition to a different node or thread.
Such an execution model, with minor variations, is adopted, for example,
by \citeN{pyop2isc}, \citeN{Fenics}, \citeN{fluidity_manual_v4},
\citeN{lizst}.

@@

We aim to maximize our impact on the platforms that are realistically used
for finite element applications, so we target conventional CPU architectures. 
The key limiting factor to the execution on GPUs is the stringent memory requirements. 
Only relatively small meshes can fit a GPU memory, and distributed execution is minimally
supported by available frameworks (if any). There has been some research on
adapting local assembly to GPUs (which we mention in Section~\ref{sec:related-work}),
although it differs from ours in several ways, including: (i) not relying on
automated code generation from a domain-specific language (explained next),
(ii) testing only very low order methods, (iii) not optimizing for
cross-loop arithmetic intensity (the goal is rather effective multi-thread
parallelization). In addition, our code transformations would drastically impact the 
GPU parallelization strategy, for example by increasing a thread's working set. 
For all these reasons, a study on extending our work to GPU architectures is 
beyond the scope of this paper. In Section~\ref{sec:generality}, however, we provide 
some intuitions about this research direction.

@@

  parallel. Throughout the paper we assumed a kernel to be executed by a single thread, which is likely to be the best strategy for standard multi-core CPUs. On the other hand, we note that for certain architectures (e.g. GPUs) this could be prohibitive due to memory requirements. Intra-kernel parallelization is one possible solution: a domain-specific compiler such as COFFEE could map mathematical quantifiers and operators to different parallelization schemes and generate distinct variants of multi-threaded kernel code. Based on our experience, we believe this is the right approach to achieve performance portability.

@@

  array sizes and loop bounds, like $n1$ and $n2$ in

>>>>

In many fields such as computational fluid dynamics, computational
electromagnetics and structural mechanics, phenomena are modelled by
partial differential equations (PDEs). Numerical techniques, such as the
finite volume method and the finite element method, are widely employed
to approximate solutions of these PDEs. Unstructured meshes are often
used to discretize the computational domain, since they allow an
accurate representation of complex geometries. The solution is sought by
applying suitable numerical operations, described by computational
kernels, to all entities of a mesh, such as edges, vertices, or cells.
On standard clusters of multicores, typically, a kernel is executed
sequentially by a thread, while parallelism is achieved by partitioning
the mesh and assigning each partition to a different node or thread.
Such an execution model, with minor variations, is adopted, for example,
by \citeN{pyop2isc}, \citeN{Fenics}, \citeN{fluidity_manual_v4},
\citeN{lizst}.

@@

We aim to maximize our impact on the platforms that are realistically used
for finite element applications, so we target conventional CPU architectures
rather than GPUs. 
The key limiting factor to the execution on GPUs is the stringent memory requirements. 
Only relatively small problems fit in a GPU memory, and support for
distributed GPU execution in general purpose finite element frameworks is minimal. There has been some research on
adapting local assembly to GPUs (which we mention in Section~\ref{sec:related-work}),
although it differs from ours in several ways, including: (i) not relying on
automated code generation from a domain-specific language (explained next),
(ii) testing only very low order methods, (iii) not optimizing for
cross-loop arithmetic intensity (the goal is rather effective multi-thread
parallelization). In addition, our code transformations would drastically impact the 
GPU parallelization strategy, for example by increasing a thread's working set. 
For all these reasons, a study on extending our work to GPU architectures is 
beyond the scope of this paper. In Section~\ref{sec:generality}, however, we provide 
some intuitions about this research direction.

@@

  parallel. Throughout the paper we assumed that a kernel will be executed
  by a single thread, which is likely to be the best strategy for standard
  multi-core CPUs. On the other hand, we note that for certain architectures
  (for example GPUs) this could be prohibitive due to memory requirements. Intra-kernel parallelization is one possible solution: a domain-specific compiler such as COFFEE could map mathematical quantifiers and operators to different parallelization schemes and generate distinct variants of multi-threaded kernel code. Based on our experience, we believe this is the right approach to achieve performance portability.

@@

  array sizes and loop bounds, for example $n1$ and $n2$ in
